---
title: Statistical analysis
subtitle: Emerald glassfrog mating activity
author: <a href="https://marce10.github.io/">Marcelo Araya-Salas</a>
date: "`r Sys.Date()`"
toc: true
toc-depth: 3
toc-location: left
number-sections: true
highlight-style: pygments
format:
  html:
    df-print: kable
    code-fold: true
    code-tools: true
    css: qmd.css
editor_options: 
  chunk_output_type: console
---

```{r set root directory}
#| eval: true
#| echo: false

# set working directory 
knitr::opts_knit$set(root.dir =  "..")

```

```{r add link to github repo}
#| eval: true
#| output: asis
#| echo: false

# print link to github repo if any
if (file.exists("./.git/config")){
  config <- readLines("./.git/config")
  url <- grep("url",  config, value = TRUE)
  url <- gsub("\\turl = |.git$", "", url)
  if (nchar(url) > 1)
  cat("\nSource code and data found at [", url, "](", url, ")", sep = "")
  }

```

```{r setup style}
#| message: false
#| warning: false

# options to customize chunk outputs
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
 )

```

<!-- skyblue box -->

::: {.alert .alert-info}
# Purpose {.unnumbered .unlisted}

-   Infer the effects of environmental variables on emerald glassfrog mating activity
:::

 


 

# Analysis flowchart {.unnumbered .unlisted}

```{mermaid}

flowchart LR
  A[Read data] --> B(Format data) 
  B --> C(Define DAGs)
  C --> D(Statistical analysis)
  D --> E(Model summary) 

style A fill:#382A5433
style B fill:#395D9C33
style C fill:#3497A933
style D fill:#60CEAC33

```

# Load packages {.unnumbered .unlisted}

```{r load packages}

# knitr is require for creating html/pdf/word reports
# formatR is used for soft-wrapping code

# install/ load packages
sketchy::load_packages(packages = c("ggplot2", "viridis", "dagitty", "ggdag", "lunar", "brms", "brmsish", "rstan"))

# set theme globally
  theme_set(theme_classic(base_size = 14))

```

# Read data

```{r}
dat <- read.csv("./data/raw/males and pairs_all years_updated.csv")

dat$org.date <- dat$date

dat$date <- paste(dat$year,  dat$date, sep = "-")

# convert "8-Jun" to month-day format
dat$date <- as.Date(dat$date, format = "%Y-%d-%b")

#convert to "1970-01-02" format
dat$date <- as.Date(dat$date, origin = "1970-01-01")

## add moon
dat$moonlight <- lunar.illumination(dat$date, shift = -6)

```

Prepare data
```{r}
# scale continuous variables
dat$mean_temp_sc <- scale(dat$mean_temp)
dat$min_temp_sc <- scale(dat$min_temp)
dat$max_temp_sc <- scale(dat$max_temp)
dat$mean_rh_sc <- scale(dat$mean_rh)
dat$total_rainfall_sc <- scale(dat$total_rainfall)
dat$prev_rainfall_sc <- scale(dat$prev_rainfall)
dat$moonlight_sc <- scale(dat$moonlight)
dat$year <- as.factor(dat$year)
dat$num_males_sc <- scale(dat$num_males)
```

# Covariance temperature

```{r}

cor(dat[, c("mean_temp", "max_temp", "min_temp", "mean_rh")])

```



# DAG + stats

## Male activity
```{r}

dag_l <- dagify(
    total_rainfall_sc ~ evapotranspiration,
    prev_rainfall_sc ~ evapotranspiration,
    mean_temp_sc ~ total_rainfall_sc,
    mean_rh_sc ~ total_rainfall_sc,
    num_males ~ mean_rh_sc,
    num_males ~ moonlight_sc,
    mean_rh_sc ~ mean_temp_sc,
    mean_rh_sc ~ prev_rainfall_sc,
    mean_rh_sc ~ total_rainfall_sc,
    num_males ~ mean_temp_sc,
    num_males ~ prev_rainfall_sc,
    num_males ~ total_rainfall_sc,
    labels = c(
        num_males = "Active\nmales",
        mean_rh_sc = "Relative\nhumidity",
        total_rainfall_sc = "Rainfall",
        prev_rainfall_sc = "Previous\nrainfall",
        moonlight_sc = "Moonlight",
        mean_temp_sc = "Temperature",
        evapotranspiration = "Evapotrans-\npiration",
        latent = c("evapotrans-\npiration"),
        outcome = "num_males"
    )
)

set.seed(5)
tidy_dag <- tidy_dagitty(dag_l)
tidy_dag$data$type <- ifelse(is.na(tidy_dag$data$to), "outcome", "predictor")
tidy_dag$data$type[tidy_dag$data$name %in% c("evapotranspiration")] <- "latent"

ggplot(tidy_dag, aes(
    x = x,
    y = y,
    xend = xend,
    yend = yend
)) + scale_color_viridis_d(
    option = "G",
    begin = 0.2,
    end = 0.8,
    alpha = 0.5
) + geom_dag_edges_fan(
    edge_color = mako(10, alpha = 0.4)[2],
    arrow = grid::arrow(length = grid::unit(10, "pt"), type = "closed")
) + geom_dag_point(aes(color = type), size = 24) + 
    geom_dag_text(color = "black", aes(label = label, color = label)) +
    theme_dag() + 
    theme(legend.position = "bottom")

```

### SEM model
```{r}
#| eval: false
#| results: 'asis'

# ----------------------------
# Submodels with year random intercept
# ----------------------------

bf_total_rain <- bf(
  total_rainfall_sc ~ 1 + (1 | year)
)

bf_prev_rain <- bf(
  prev_rainfall_sc ~ 1 + (1 | year)
)

bf_temp <- bf(
  temp_sc ~ 1 + total_rainfall_sc + (1 | year)
)

bf_rh <- bf(
  mean_rh_sc ~ 1 +
    temp_sc +
    prev_rainfall_sc +
    total_rainfall_sc +
    (1 | year)
)

bf_males <- bf(
  num_males ~ 
    s(mean_rh_sc) +
    temp_sc +
    prev_rainfall_sc +
    total_rainfall_sc +
    moonlight_sc +
    (1 | year),
  family = negbinomial()
)



# ----------------------------
# Weakly informative priors
# ----------------------------

priors <- c(

  # ----------------------------
  # Linear slopes (all z-scored)
  # ----------------------------

  prior(normal(0, 0.5), class = "b", resp = "meanrhsc"),
  prior(normal(0, 0.5), class = "b", resp = "tempsc"),
  prior(normal(0, 0.5), class = "b", resp = "nummales"),

  # ----------------------------
  # Smooth term (humidity → males)
  # ----------------------------

  prior(normal(0, 1), class = "sds", resp = "nummales"),

  # ----------------------------
  # Year random effects
  # ----------------------------

  # ----------------------------
  # Negative binomial dispersion
  # ----------------------------

  prior(exponential(1), class = "shape", resp = "nummales")
)

dat$temp_sc <- dat$mean_temp_sc

# run a prior sampling model to check the priors
sem_prior_only <- brm(
  bf_total_rain +
    bf_prev_rain +
    bf_temp +
    bf_rh +
    bf_males +
    set_rescor(FALSE),
  data = dat,
  prior = priors,
  sample_prior = "only",
  iter = 2000,
  chains = 2,
  cores = 2,
  # backend = "cmdstanr",
  file = "male_sem_prior_only.RDS",
  file_refit = "on_change")

# check there are not flat priors
ps <- prior_summary(sem_prior_only)

any(grepl("flat", ps$prior))


# ----------------------------
# Fit full SEM
# ----------------------------

sem_model <- brm(
  bf_total_rain +
    bf_prev_rain +
    bf_temp +
    bf_rh +
    bf_males +
    set_rescor(FALSE),
  data = dat,
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  control = list(adapt_delta = 0.95),
    file = "./data/processed/male_sem_model.RDS",
    backend = "cmdstanr",
  file_refit = "on_change"
)

add_criterion(sem_model, criterion = c("loo"))

dat$temp_sc <- dat$max_temp_sc

sem_model_max_temp <- brm(
  bf_total_rain +
    bf_prev_rain +
    bf_temp +
    bf_rh +
    bf_males +
    set_rescor(FALSE),
  data = dat,
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  control = list(adapt_delta = 0.95),
    file = "./data/processed/male_sem_model_max_temp.RDS",
    backend = "cmdstanr",
  file_refit = "on_change"
)

add_criterion(sem_model_max_temp, criterion = c("loo"))


dat$temp_sc <- dat$min_temp_sc


sem_model_min_temp <- brm(
  bf_total_rain +
    bf_prev_rain +
    bf_temp +
    bf_rh +
    bf_males +
    set_rescor(FALSE),
  data = dat,
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  control = list(adapt_delta = 0.95),
    file = "./data/processed/male_sem_model_min_temp.RDS",
    backend = "cmdstanr",
  file_refit = "on_change"
)

add_criterion(sem_model_min_temp, criterion = c("loo"))

```

### Model selection
```{r}

mean_temp_males_mod <- readRDS("./data/processed/male_sem_model.RDS")
max_temp <- readRDS("./data/processed/male_sem_model_max_temp.RDS")
min_temp <- readRDS("./data/processed/male_sem_model_min_temp.RDS")


loo_diff <- loo::loo_compare(mean_temp_males_mod, max_temp, min_temp)

rows <- rownames(loo_diff)
loo_diff <- as.data.frame(loo_diff)

loo_diff$model <- rows

loo_diff

```

#### Best model
```{r}
#| results: asis

extended_summary(mean_temp_males_mod, highlight = TRUE, trace.palette = mako)

```

#### Evaluate "year" effect
```{r}
#| eval: false

# ----------------------------
# Submodels with year random intercept
# ----------------------------

bf_total_rain <- bf(
  total_rainfall_sc ~ 1
)

bf_prev_rain <- bf(
  prev_rainfall_sc ~ 1
)

bf_temp <- bf(
  temp_sc ~ 1 + total_rainfall_sc
)

bf_rh <- bf(
  mean_rh_sc ~ 1 +
    temp_sc +
    prev_rainfall_sc +
    total_rainfall_sc
)

bf_males <- bf(
  num_males ~ 
    s(mean_rh_sc) +
    temp_sc +
    prev_rainfall_sc +
    total_rainfall_sc +
    moonlight_sc,
  family = negbinomial()
)



# ----------------------------
# Weakly informative priors
# ----------------------------

priors <- c(

  # ----------------------------
  # Linear slopes (all z-scored)
  # ----------------------------

  prior(normal(0, 0.5), class = "b", resp = "meanrhsc"),
  prior(normal(0, 0.5), class = "b", resp = "tempsc"),
  prior(normal(0, 0.5), class = "b", resp = "nummales"),

  # ----------------------------
  # Smooth term (humidity → males)
  # ----------------------------

  prior(normal(0, 1), class = "sds", resp = "nummales"),

  # ----------------------------
  # Year random effects
  # ----------------------------

  # ----------------------------
  # Negative binomial dispersion
  # ----------------------------

  prior(exponential(1), class = "shape", resp = "nummales")
)

dat$temp_sc <- dat$mean_temp_sc



# ----------------------------
# Fit full SEM
# ----------------------------

sem_model_no_year <- brm(
  bf_total_rain +
    bf_prev_rain +
    bf_temp +
    bf_rh +
    bf_males +
    set_rescor(FALSE),
  data = dat,
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  control = list(adapt_delta = 0.95),
    file = "./data/processed/male_sem_model_no_year.RDS",
    backend = "cmdstanr",
  file_refit = "on_change"
)

add_criterion(sem_model_no_year, criterion = c("loo"))

```

```{r}
no_year <- readRDS("./data/processed/male_sem_model_no_year.RDS")

loo_diff <- loo::loo_compare(mean_temp_males_mod, no_year)

rows <- rownames(loo_diff)
loo_diff <- as.data.frame(loo_diff)

loo_diff$model <- rows

loo_diff

R2_full <- bayes_R2(mean_temp_males_mod, resp = "nummales",  summary = FALSE)

R2_fixed <- bayes_R2(
  mean_temp_males_mod,
  resp = "nummales",
  re_formula = NA,
   summary = FALSE
)

str(R2_full)
    
delta_R2 <- R2_full - R2_fixed

quantile(delta_R2, c(0.025, 0.5, 0.975))

```

### Graph raw data

Temperature
```{r}
temp_grid <- data.frame(
  mean_temp = seq(min(dat$mean_temp, na.rm = TRUE),
                  max(dat$mean_temp, na.rm = TRUE),
                  length.out = 100)
)
temp_grid$mean_rh_sc        <- 0
temp_grid$prev_rainfall_sc <- 0
temp_grid$total_rainfall_sc <- 0
temp_grid$moonlight_sc     <- 0


temp_center <- attr(dat$mean_temp_sc, "scaled:center")
temp_scale  <- attr(dat$mean_temp_sc, "scaled:scale")

temp_grid$temp_sc <- (temp_grid$mean_temp - temp_center) / temp_scale


pred <- fitted(
  mean_temp_males_mod,
  resp = "nummales",
  newdata = temp_grid,
  re_formula = NA,   # population-level effect
  summary = TRUE
)

temp_grid <- cbind(temp_grid, pred)


ggplot(data = dat, aes(x = mean_temp, y = num_males)) +
    geom_point(alpha = 0.4, color = mako(6, alpha = 0.6)[3], size = 3) +
  geom_line(data = temp_grid,
            aes(x = mean_temp, y = Estimate),
            linewidth = 1.2,
            color = mako(10, alpha = 0.8)[3]) +
  geom_ribbon(data = temp_grid,
              aes(x = mean_temp, y = Estimate, ymin = Q2.5, ymax = Q97.5),
              alpha = 0.1,
              fill = "black") +
       labs(x = "Mean temperature (°C)", y = "Number of males")

# by year
temp_grid_year <- data.frame(
  mean_temp = seq(min(dat$mean_temp, na.rm = TRUE),
                  max(dat$mean_temp, na.rm = TRUE),
                  length.out = 100),
  year = unique(dat$year)
)
temp_grid_year$mean_rh_sc        <- 0
temp_grid_year$prev_rainfall_sc <- 0
temp_grid_year$total_rainfall_sc <- 0
temp_grid_year$moonlight_sc     <- 0


temp_center <- attr(dat$mean_temp_sc, "scaled:center")
temp_scale  <- attr(dat$mean_temp_sc, "scaled:scale")

temp_grid_year$temp_sc <- (temp_grid_year$mean_temp - temp_center) / temp_scale


pred <- fitted(
  mean_temp_males_mod,
  resp = "nummales",
  newdata = temp_grid_year,
  re_formula = NULL,   # <-- this is the key
  summary = TRUE
)

temp_grid_year <- cbind(temp_grid_year, pred)

ggplot(data = dat, aes(x = mean_temp, y = num_males)) +
    geom_point(alpha = 0.4, color = mako(6, alpha = 0.6)[3], size = 3) +
  geom_line(data = temp_grid_year,
            aes(x = mean_temp, y = Estimate),
            linewidth = 1.2,
            color = mako(10, alpha = 0.8)[3]) +
    facet_wrap(~year) +
  geom_ribbon(data = temp_grid_year,
              aes(x = mean_temp, y = Estimate, ymin = Q2.5, ymax = Q97.5),
              alpha = 0.1,
              fill = "black") +
    labs(x = "Mean temperature (°C)", y = "Number of males")
```

Previous rainfall
```{r}

ggplot(data = dat, aes(x = prev_rainfall, y = num_males)) +
    geom_point(alpha = 0.4, color = mako(6, alpha = 0.6)[3], size = 3) +
       labs(x = "Previous rainfall (mm)", y = "Number of males")

```

Total rainfall
```{r}

ggplot(data = dat, aes(x = total_rainfall, y = num_males)) +
    geom_point(alpha = 0.4, color = mako(6, alpha = 0.6)[3], size = 3) +
       labs(x = "Rainfall (mm)", y = "Number of males")

```

Moonlight
```{r}

ggplot(data = dat, aes(x = moonlight * 100, y = num_males)) +
    geom_point(alpha = 0.4, color = mako(6, alpha = 0.6)[3], size = 3) +
       labs(x = "Moonlight (%)", y = "Number of males")

 
```


## Female activity
```{r}

dag_l <- dagify(
    num_males_sc ~ mean_rh_sc,
    num_males_sc ~ moonlight_sc,
    num_males_sc ~ mean_temp_sc,
    num_males_sc ~ prev_rainfall_sc,
    num_males_sc ~ total_rainfall_sc,
    total_rainfall_sc ~ evapotranspiration,
    prev_rainfall_sc ~ evapotranspiration,
    mean_temp_sc ~ total_rainfall_sc,
    mean_rh_sc ~ total_rainfall_sc,
    num_pairs ~ mean_rh_sc,
    num_pairs ~ moonlight_sc,
    mean_rh_sc ~ mean_temp_sc,
    mean_rh_sc ~ prev_rainfall_sc,
    mean_rh_sc ~ total_rainfall_sc,
    num_pairs ~ mean_temp_sc,
    num_pairs ~ prev_rainfall_sc,
    num_pairs ~ total_rainfall_sc,
    num_pairs ~ num_males_sc,
    labels = c(
        num_pairs = "Active\nfemales",
        mean_rh_sc = "Relative\nhumidity",
        total_rainfall_sc = "Rainfall",
        prev_rainfall_sc = "Previous\nrainfall",
        moonlight_sc = "Moonlight",
        mean_temp_sc = "Temperature",
        num_males_sc = "Active\nmales",
        evapotranspiration = "Evapotrans-\npiration",
        latent = c("evapotrans-\npiration"),
        outcome = "num_pairs"
    )
)

set.seed(5)
tidy_dag <- tidy_dagitty(dag_l)
tidy_dag$data$type <- ifelse(is.na(tidy_dag$data$to), "outcome", "predictor")
tidy_dag$data$type[tidy_dag$data$name %in% c("evapotranspiration")] <- "latent"

ggplot(tidy_dag, aes(
    x = x,
    y = y,
    xend = xend,
    yend = yend
)) + scale_color_viridis_d(
    option = "G",
    begin = 0.2,
    end = 0.8,
    alpha = 0.5
) + geom_dag_edges_fan(
    edge_color = mako(10, alpha = 0.4)[2],
    arrow = grid::arrow(length = grid::unit(10, "pt"), type = "closed")
) + geom_dag_point(aes(color = type), size = 24) + 
    geom_dag_text(color = "black", aes(label = label, color = label)) +
    theme_dag() + 
    theme(legend.position = "bottom")

```

### SEM model
```{r}
#| eval: false
#| results: 'asis'

# ----------------------------
# Submodels
# ----------------------------

bf_total_rain <- bf(
  total_rainfall_sc ~ 1 + (1 | year)
)

bf_prev_rain <- bf(
  prev_rainfall_sc ~ 1 + (1 | year)
)

bf_temp <- bf(
  temp_sc ~ 1 +
    total_rainfall_sc +
    (1 | year)
)

bf_rh <- bf(
  mean_rh_sc ~ 1 +
    temp_sc +
    prev_rainfall_sc +
    total_rainfall_sc +
    (1 | year)
)

# mediator: active males (count)
bf_males <- bf(
  num_males ~ 
    s(mean_rh_sc) +
    temp_sc +
    prev_rainfall_sc +
    total_rainfall_sc +
    moonlight_sc +
    (1 | year),
  family = negbinomial()
)

# outcome: active females (count)
bf_pairs <- bf(
  num_pairs ~ 
    s(mean_rh_sc) +
    temp_sc +
    prev_rainfall_sc +
    total_rainfall_sc +
    moonlight_sc +
    num_males_sc +
    (1 | year),
  family = negbinomial()
)

# ----------------------------
# Weakly informative priors
# ----------------------------

priors <- c(

  # Linear slopes (all predictors z-scored)
  prior(normal(0, 0.5), class = "b", resp = "meanrhsc"),
  prior(normal(0, 0.5), class = "b", resp = "tempsc"),
  prior(normal(0, 0.5), class = "b", resp = "nummales"),
  prior(normal(0, 0.5), class = "b", resp = "numpairs"),

  # Smooth terms (humidity effects)
  prior(normal(0, 1), class = "sds", resp = "nummales"),
  prior(normal(0, 1), class = "sds", resp = "numpairs"),

  # Negative binomial dispersion
  prior(exponential(1), class = "shape", resp = "nummales"),
  prior(exponential(1), class = "shape", resp = "numpairs")
)

# ----------------------------
# Prior predictive check
# ----------------------------

sem_prior_only <- brm(
  bf_total_rain +
    bf_prev_rain +
    bf_temp +
    bf_rh +
    bf_males +
    bf_pairs +
    set_rescor(FALSE),
  data = dat,
  prior = priors,
  sample_prior = "only",
  iter = 2000,
  chains = 2,
  cores = 2,
  backend = "cmdstanr",
    file = "female_sem_prior_only.RDS",
  file_refit = "on_change"
)

ps <- prior_summary(sem_prior_only)

any(grepl("flat", ps$prior))


# ----------------------------
# Fit full SEM
# ----------------------------

fem_sem_model <- brm(
  bf_total_rain +
    bf_prev_rain +
    bf_temp +
    bf_rh +
    bf_males +
    bf_pairs +
    set_rescor(FALSE),
  data = dat,
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  control = list(adapt_delta = 0.95),
  file = "./data/processed/female_sem_model.RDS",
  backend = "cmdstanr",
  file_refit = "on_change"
)

add_criterion(fem_sem_model, criterion = c("loo"))

dat$temp_sc <- dat$max_temp_sc

fem_max_temp <- brm(
  bf_total_rain +
    bf_prev_rain +
    bf_temp +
    bf_rh +
    bf_males +
    bf_pairs +
    set_rescor(FALSE),
  data = dat,
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  control = list(adapt_delta = 0.95),
  file = "./data/processed/female_sem_model_max_temp.RDS",
  backend = "cmdstanr",
  file_refit = "on_change"
)

add_criterion(fem_max_temp, criterion = c("loo"))

dat$temp_sc <- dat$min_temp_sc

fem_min_temp <- brm(
  bf_total_rain +
    bf_prev_rain +
    bf_temp +
    bf_rh +
    bf_males +
    bf_pairs +
    set_rescor(FALSE),
  data = dat,
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  control = list(adapt_delta = 0.95),
  file = "./data/processed/female_sem_model_min_temp.RDS",
  backend = "cmdstanr",
  file_refit = "on_change"
)

add_criterion(fem_min_temp, criterion = c("loo"))

```

### Model selection
```{r}

mean_temp_females_mod <- readRDS("./data/processed/female_sem_model.RDS")
max_temp <- readRDS("./data/processed/female_sem_model_max_temp.RDS")
min_temp <- readRDS("./data/processed/female_sem_model_min_temp.RDS")

loo_diff <- loo::loo_compare(mean_temp_females_mod, max_temp, min_temp)

rows <- rownames(loo_diff)
loo_diff <- as.data.frame(loo_diff)

loo_diff$model <- rows

loo_diff

```

#### Best model
```{r}
#| results: asis

extended_summary(
  mean_temp_females_mod,
  highlight = TRUE, 
  trace.palette = mako
)


```

#### Evaluate "year" effect
```{r}
#| eval: false

# ----------------------------
# Submodels with year random intercept
# ----------------------------


bf_total_rain <- bf(
  total_rainfall_sc ~ 1 
)

bf_prev_rain <- bf(
  prev_rainfall_sc ~ 1 
)

bf_temp <- bf(
  temp_sc ~ 1 +
    total_rainfall_sc
)

bf_rh <- bf(
  mean_rh_sc ~ 1 +
    temp_sc +
    prev_rainfall_sc +
    total_rainfall_sc
)

# mediator: active males (count)
bf_males <- bf(
  num_males ~ 
    s(mean_rh_sc) +
    temp_sc +
    prev_rainfall_sc +
    total_rainfall_sc +
    moonlight_sc,
  family = negbinomial()
)

# outcome: active females (count)
bf_pairs <- bf(
  num_pairs ~ 
    s(mean_rh_sc) +
    temp_sc +
    prev_rainfall_sc +
    total_rainfall_sc +
    moonlight_sc +
    num_males_sc,
  family = negbinomial()
)

# ----------------------------
# Weakly informative priors
# ----------------------------

priors <- c(

  # Linear slopes (all predictors z-scored)
  prior(normal(0, 0.5), class = "b", resp = "meanrhsc"),
  prior(normal(0, 0.5), class = "b", resp = "tempsc"),
  prior(normal(0, 0.5), class = "b", resp = "nummales"),
  prior(normal(0, 0.5), class = "b", resp = "numpairs"),

  # Smooth terms (humidity effects)
  prior(normal(0, 1), class = "sds", resp = "nummales"),
  prior(normal(0, 1), class = "sds", resp = "numpairs"),

  # Negative binomial dispersion
  prior(exponential(1), class = "shape", resp = "nummales"),
  prior(exponential(1), class = "shape", resp = "numpairs")
)

# Fit full SEM
fem_sem_model_no_year <- brm(
  bf_total_rain +
    bf_prev_rain +
    bf_temp +
    bf_rh +
    bf_males +
    bf_pairs +
    set_rescor(FALSE),
  data = dat,
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  control = list(adapt_delta = 0.95),
  file = "./data/processed/female_sem_model_no_year.RDS",
  backend = "cmdstanr",
  file_refit = "on_change"
)

add_criterion(fem_sem_model_no_year, criterion = c("loo"))


```

```{r}
no_year <- readRDS("./data/processed/female_sem_model_no_year.RDS")

loo_diff <- loo::loo_compare(mean_temp_females_mod, no_year)

rows <- rownames(loo_diff)
loo_diff <- as.data.frame(loo_diff)

loo_diff$model <- rows

loo_diff

R2_full <- bayes_R2(mean_temp_females_mod, resp = "numpairs",  summary = FALSE)

R2_fixed <- bayes_R2(
  mean_temp_females_mod,
  resp = "numpairs",
  re_formula = NA,
   summary = FALSE
)

str(R2_full)
    
delta_R2 <- R2_full - R2_fixed

quantile(delta_R2, c(0.025, 0.5, 0.975))

```

### Graph raw data

Temperature
```{r}

temp_grid <- data.frame(
  mean_temp = seq(min(dat$mean_temp, na.rm = TRUE),
                  max(dat$mean_temp, na.rm = TRUE),
                  length.out = 100)
)
temp_grid$mean_rh_sc        <- 0
temp_grid$prev_rainfall_sc <- 0
temp_grid$total_rainfall_sc <- 0
temp_grid$moonlight_sc     <- 0
temp_grid$num_males_sc     <- 0


temp_center <- attr(dat$mean_temp_sc, "scaled:center")
temp_scale  <- attr(dat$mean_temp_sc, "scaled:scale")

temp_grid$temp_sc <- (temp_grid$mean_temp - temp_center) / temp_scale


pred <- fitted(
  mean_temp_females_mod,
  resp = "numpairs",
  newdata = temp_grid,
  re_formula = NA,   # population-level effect
  summary = TRUE
)

temp_grid <- cbind(temp_grid, pred)

ggplot(data = dat, aes(x = mean_temp, y = num_pairs)) +
    geom_point(alpha = 0.4, color = mako(6, alpha = 0.6)[3], size = 3)  +
       labs(x = "Mean temperature (°C)", y = "Number of amplexus")

ggplot(data = dat, aes(x = mean_temp, y = num_pairs)) +
    geom_point(alpha = 0.4, color = mako(6, alpha = 0.6)[3], size = 3) +
  geom_line(data = temp_grid,
            aes(x = mean_temp, y = Estimate),
            linewidth = 1.2,
            color = mako(10, alpha = 0.8)[3]) +
  geom_ribbon(data = temp_grid,
              aes(x = mean_temp, y = Estimate, ymin = Q2.5, ymax = Q97.5),
              alpha = 0.1,
              fill = "black") +
       labs(x = "Mean temperature (°C)", y = "Number of amplexus")

# by year
temp_grid_year <- data.frame(
  mean_temp = seq(min(dat$mean_temp, na.rm = TRUE),
                  max(dat$mean_temp, na.rm = TRUE),
                  length.out = 100),
  year = unique(dat$year)
)
temp_grid_year$mean_rh_sc        <- 0
temp_grid_year$prev_rainfall_sc <- 0
temp_grid_year$total_rainfall_sc <- 0
temp_grid_year$moonlight_sc     <- 0
temp_grid_year$num_males_sc     <- 0

temp_center <- attr(dat$mean_temp_sc, "scaled:center")
temp_scale  <- attr(dat$mean_temp_sc, "scaled:scale")

temp_grid_year$temp_sc <- (temp_grid_year$mean_temp - temp_center) / temp_scale


pred <- fitted(
  mean_temp_females_mod,
  resp = "numpairs",
  newdata = temp_grid_year,
  re_formula = NULL,   # <-- this is the key
  summary = TRUE
)

temp_grid_year <- cbind(temp_grid_year, pred)

ggplot(data = dat, aes(x = mean_temp, y = num_pairs)) +
    geom_point(alpha = 0.4, color = mako(6, alpha = 0.6)[3], size = 3) +
    facet_wrap(~year) +
    labs(x = "Mean temperature (°C)", y = "Number of amplexus")

ggplot(data = dat, aes(x = mean_temp, y = num_pairs)) +
    geom_point(alpha = 0.4, color = mako(6, alpha = 0.6)[3], size = 3) +
  geom_line(data = temp_grid_year,
            aes(x = mean_temp, y = Estimate),
            linewidth = 1.2,
            color = mako(10, alpha = 0.8)[3]) +
    facet_wrap(~year) +
  geom_ribbon(data = temp_grid_year,
              aes(x = mean_temp, y = Estimate, ymin = Q2.5, ymax = Q97.5),
              alpha = 0.1,
              fill = "black") +
    labs(x = "Mean temperature (°C)", y = "Number of amplexus")

  
```

Total rainfall
```{r}

temp_grid <- data.frame(
  total_rainfall = seq(min(dat$total_rainfall, na.rm = TRUE),
                  max(dat$total_rainfall, na.rm = TRUE),
                  length.out = 100)
)

temp_grid$temp_sc     <- 0
temp_grid$mean_rh_sc        <- 0
temp_grid$prev_rainfall_sc <- 0
temp_grid$moonlight_sc     <- 0
temp_grid$num_males_sc     <- 0


temp_center <- attr(dat$total_rainfall_sc, "scaled:center")
temp_scale  <- attr(dat$total_rainfall_sc, "scaled:scale")

temp_grid$total_rainfall_sc <- (temp_grid$total_rainfall - temp_center) / temp_scale


pred <- fitted(
  mean_temp_females_mod,
  resp = "numpairs",
  newdata = temp_grid,
  re_formula = NA,   # population-level effect
  summary = TRUE
)

temp_grid <- cbind(temp_grid, pred)


ggplot(data = dat, aes(x = total_rainfall, y = num_pairs)) +
    geom_point(alpha = 0.4, color = mako(6, alpha = 0.6)[3], size = 3) +
  geom_line(data = temp_grid,
            aes(x = total_rainfall, y = Estimate),
            linewidth = 1.2,
            color = mako(10, alpha = 0.8)[3]) +
  geom_ribbon(data = temp_grid,
              aes(x = total_rainfall, y = Estimate, ymin = Q2.5, ymax = Q97.5),
              alpha = 0.1,
              fill = "black") +
       labs(x = "Maximum rainfall (mm)", y = "Number of amplexus")


# by year
temp_grid_year <- expand.grid(
  total_rainfall = seq(min(dat$total_rainfall, na.rm = TRUE),
                  max(dat$total_rainfall, na.rm = TRUE)),
                  length.out = 100,
  year = unique(dat$year))

temp_grid_year$temp_sc     <- 0
temp_grid_year$mean_rh_sc        <- 0
temp_grid_year$prev_rainfall_sc <- 0
temp_grid_year$moonlight_sc     <- 0
temp_grid_year$num_males_sc     <- 0

temp_center <- attr(dat$total_rainfall_sc, "scaled:center")
temp_scale  <- attr(dat$total_rainfall_sc, "scaled:scale")

temp_grid_year$total_rainfall_sc <- (temp_grid_year$total_rainfall - temp_center) / temp_scale


pred <- fitted(
  mean_temp_females_mod,
  resp = "numpairs",
  newdata = temp_grid_year,
  re_formula = NULL,   # <-- this is the key
  summary = TRUE
)

temp_grid_year <- cbind(temp_grid_year, pred)

ggplot(data = dat, aes(x = total_rainfall, y = num_pairs)) +
    geom_point(alpha = 0.4, color = mako(6, alpha = 0.6)[3], size = 3) +
  geom_line(data = temp_grid_year,
            aes(x = total_rainfall, y = Estimate),
            linewidth = 1.2,
            color = mako(10, alpha = 0.8)[3]) +
    facet_wrap(~year) +
  geom_ribbon(data = temp_grid_year,
              aes(x = total_rainfall, y = Estimate, ymin = Q2.5, ymax = Q97.5),
              alpha = 0.1,
              fill = "black") +
    labs(x = "Maximum rainfall (mm)", y = "Number of amplexus")

  
```

Number of males
```{r}

temp_grid <- data.frame(
  num_males = seq(min(dat$num_males, na.rm = TRUE),
                  max(dat$num_males, na.rm = TRUE),
                  length.out = 100)
)

temp_grid$temp_sc     <- 0
temp_grid$mean_rh_sc        <- 0
temp_grid$prev_rainfall_sc <- 0
temp_grid$total_rainfall_sc <- 0
temp_grid$moonlight_sc     <- 0


temp_center <- attr(dat$num_males_sc, "scaled:center")
temp_scale  <- attr(dat$num_males_sc, "scaled:scale")

temp_grid$num_males_sc <- (temp_grid$num_males - temp_center) / temp_scale


pred <- fitted(
  mean_temp_females_mod,
  resp = "numpairs",
  newdata = temp_grid,
  re_formula = NA,   # population-level effect
  summary = TRUE
)

temp_grid <- cbind(temp_grid, pred)

ggplot(data = dat, aes(x = num_males, y = num_pairs)) +
    geom_point(alpha = 0.4, color = mako(6, alpha = 0.6)[3], size = 3) +
  geom_line(data = temp_grid,
            aes(x = num_males, y = Estimate),
            linewidth = 1.2,
            color = mako(10, alpha = 0.8)[3]) +
  geom_ribbon(data = temp_grid,
              aes(x = num_males, y = Estimate, ymin = Q2.5, ymax = Q97.5),
              alpha = 0.1,
              fill = "black") +
       labs(x = "Number of males", y = "Number of amplexus")


# by year
temp_grid_year <- expand.grid(
  num_males = seq(min(dat$num_males, na.rm = TRUE),
                  max(dat$num_males, na.rm = TRUE)),
                  length.out = 100,
  year = unique(dat$year))

temp_grid_year$temp_sc     <- 0
temp_grid_year$mean_rh_sc        <- 0
temp_grid_year$prev_rainfall_sc <- 0
temp_grid_year$total_rainfall_sc <- 0
temp_grid_year$moonlight_sc     <- 0

temp_center <- attr(dat$num_males_sc, "scaled:center")
temp_scale  <- attr(dat$num_males_sc, "scaled:scale")

temp_grid_year$num_males_sc <- (temp_grid_year$num_males - temp_center) / temp_scale


pred <- fitted(
  mean_temp_females_mod,
  resp = "numpairs",
  newdata = temp_grid_year,
  re_formula = NULL,   # <-- this is the key
  summary = TRUE
)

temp_grid_year <- cbind(temp_grid_year, pred)

ggplot(data = dat, aes(x = num_males, y = num_pairs)) +
    geom_point(alpha = 0.4, color = mako(6, alpha = 0.6)[3], size = 3) +
  geom_line(data = temp_grid_year,
            aes(x = num_males, y = Estimate),
            linewidth = 1.2,
            color = mako(10, alpha = 0.8)[3]) +
    facet_wrap(~year) +
  geom_ribbon(data = temp_grid_year,
              aes(x = num_males, y = Estimate, ymin = Q2.5, ymax = Q97.5),
              alpha = 0.1,
              fill = "black") +
    labs(x = "Number of males", y = "Number of amplexus")

  
```

Previous rainfall
```{r}
ggplot(data = dat, aes(x = prev_rainfall, y = num_pairs)) +
    geom_point(alpha = 0.4, color = mako(6, alpha = 0.6)[3], size = 3) +
       labs(x = "Previous rainfall (mm)", y = "Number of amplexus")
```

Moonlight
```{r}
ggplot(data = dat, aes(x = moonlight * 100, y = num_pairs)) +
    geom_point(alpha = 0.4, color = mako(6, alpha = 0.6)[3], size = 3) +
       labs(x = "Moonlight (%)", y = "Number of amplexus")
```


::: {.alert .alert-success}
# Takeaways {#takeaways .unnumbered .unlisted}

:::


# Session information {.unnumbered .unlisted}

<details>
  <summary>Click to see</summary>
```{r session info}
#| echo: false

# if devtools is installed use devtools::session_info()
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::session_info()
} else {
  sessionInfo()
}

```
</details>
